---
title: "Final_Project_ah4028.rmd"
author: "Austin Hsin"
date: "2023-12-06"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
```

# NYC Restaurant Inspections

Cleaning and organizing NYC Restaurant Inspections by violation 04K - evidence of rats

```{r inspections_df}
#Importing NYC Restaurant Inspections csv
inspections_df = read_csv("./data/NYC_Restaurant_Inspections.csv") |>
  janitor::clean_names() |>
  separate(inspection_date, c("Month", "Day", "Year"), sep = "/") |>
  select(-Day) |> #can remove other columns as needed
  filter(as.numeric(Year) > 2015)
```

Plotting 04K violations

```{r violation plot}
#Observations of violations
violations_plot_df <- inspections_df |>
  group_by(Year, boro, violation_code) |>
  mutate(code_obs = n())

#Plotting 04K violations by coordinates
violations_plot_df |>
  filter(latitude > 0) |>
  filter(violation_code == "04K") |>
  ggplot(aes(x = longitude, y = latitude, color = boro)) +
  geom_point(alpha = 0.5) +
  coord_cartesian()

#Plotting 04K violations by year and boro
violations_plot_df |>
  filter(violation_code == "04K") |>
  ggplot(aes(x = as.numeric(Year), y = code_obs, color = boro)) +
  geom_point() +
  geom_line() +
  scale_x_continuous(
    breaks = c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023), 
    labels = c(2016, 2017, 2018, 2019, 2020, 2021, 2022, 2023))

#Plotting 04K violations to other 04 violations
violations_plot_df |>
  filter(violation_code %in% c("04K","04M","04L", "04O")) |>
  ggplot(aes(x = violation_code), color = violation_code) +
  geom_bar()

#Plotting 04K violations by cuisine type
violations_plot_df |>
  filter(violation_code == "04K") |>
  group_by(cuisine_description) |>
  mutate(cuisine_obs = n()) |>
  filter(cuisine_obs > 25) |>
  ggplot(aes(x = cuisine_description, y = cuisine_obs)) +
  geom_point() +
  theme(axis.text.x = element_text(angle = 40, hjust = 1))
```

Note: consider year comparisons of geographic data? 2020-2023 from pandemic onwards
Discuss why the data is so low prior to 2020 and shot up after pandemic
Descriptive discussion on comparing to other creature violations and why that may be - whether rats are as important if they're not as prevalent
ORDER cuisine type graph + describe the graph