---
title: "Rat Sightings Analysis"
output: 
  html_document:
    code_folding: hide
    toc: TRUE
    toc_float: TRUE
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
library(tidyverse)
library(plotly)
library(lubridate)
library(dplyr)
```

```{r rat sightings, include = FALSE}
sightings = read_csv('data/NYC_Rat_Sightings.csv') |>
  janitor::clean_names() |>
  separate(created_date, into=c("month","e", "day","f", "year", "g", "time"), sep=c(2,3,5,6,10,11)) |>
  select(-e,-f,-g) |>
  mutate(date = paste(year, month, day, sep=""),
         date = as.numeric(date)) |> 
  filter(date <= 20231031,
         date >= 20160101,
         !incident_zip <= 10000,
         !incident_zip > 11697,
         !borough %in% c("Unspecified", NA)) |>
  select(-agency, -agency_name, -complaint_type, -descriptor, -landmark, -facility_type, -park_facility_name, -vehicle_type, -taxi_company_borough, -taxi_pick_up_location, -bridge_highway_name, -road_ramp, -bridge_highway_segment, -bridge_highway_direction) |> select(unique_key, date, year, month, day, everything())
```

```{r, include=FALSE, message = FALSE}
weather_df = rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2016-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

```{r, include = FALSE}
sightings$date <- as.Date(as.character(sightings$date), format = "%Y%m%d")

rat_weather =
  right_join(sightings, weather_df, by="date")
```

## Descriptive Analyses

Let's first see how the number of rat sightings in NYC have changed over time. We can see that sightings went up drastically since 2020, when the rat populations were able to thrive as people were quarantined inside.

```{r}
sightings |> group_by(year) |>
  summarize(count=n()) |>
  plot_ly(x=~year, y=~count, type="scatter", mode="line") |> layout(title = "Rat Sightings over Time", xaxis=list(title="Year"), yaxis=list(title="Number of Rat Sightings"))
```

And if we look by time of year, we find that in each year May through October have the highest count of rat sightings.

```{r, message = FALSE}
sightings |> 
  group_by(year,month) |> 
  summarize(count = n()) |> 
  ggplot(aes(x=month, y=count, color=year, group=year)) + geom_line() 
```


A time-series shows that this is an annual pattern and that the number of rat sightings has increased over time. 

```{r, message = FALSE, warning = FALSE}
rat_weather |>
  mutate(year = as.factor(year)) |>
  group_by(date) |>
  summarize(count = n(), year) |>  
  ggplot(aes(x = date, y = count, color=year)) +
  geom_point(alpha = 0.25) + 
  labs(title = "Time Series: Rat Sightings Over Time",
       x = "Date",
       y = "Number of Rat Sightings")
```

Let's further break down these trends by borough:

```{r, message = FALSE}
sightings |> 
  group_by(year, borough) |> 
  summarize(count=n()) |> 
  plot_ly(x=~year, y=~count, color=~borough, type="scatter", mode="line") |> layout(title = "Rat Sightings by Borough", xaxis=list(title="Year"), yaxis=list(title="Number of Rat Sightings"))
```

There is a drastic increase in number of rat sightings after 2020 across all boroughs except Staten Island which started to decrease this year. Brooklyn and Manhattan had the most cases overall. However, this data only covers rat sightings until October 31, 2023, so this is not representative of the entire year of 2023.

Lastly, looking at relative density by longitude and latitude:

```{r, message = FALSE, warning = FALSE}
rat_weather |>
  sample_n(10000) |>
  plot_ly(x = ~longitude, y = ~latitude, type = "scatter", mode = "markers", marker=list(size=2, opacity=0.25),
        color = ~borough, colors = c("blue", "red", "green", "purple", "orange")) |> 
  layout(title = "Rat Sightings by Latitude and Longitude",
         xaxis = list(title = "Longitude"),
         yaxis = list(title = "Latitude"))
```

Areas with more rat sightings appear to be parts of upper Manhattan, west Brooklyn, and the Bronx.

## Statistical Analyses

In April 2023, a "Rat Czar" was appointed to tackle the problem of rat presence across the city. Let's map the number of rat sightings and the linear regression in the 6 months pre vs. post-appointment.

```{r, message = FALSE, echo=FALSE}
# looking at rat czar effects
rat_czar = sightings |>
   mutate(date = paste(year, month, day, sep=""),
         date = as.numeric(date)) |> 
  filter(date >= 20221001,
         !incident_zip <= 10000,
         !incident_zip > 11697,
         !borough %in% c("Unspecified", NA)) |>
  group_by(year, month) |>
  summarize(count=n()) |>
  mutate(month_year = paste(month, year, sep="_"),
         month_year = as.factor(month_year)) 
```

```{r prepost counts}
# graph of counts per month, 6 months before and after coronation
rat_czar |>
  ggplot(aes(x=month_year, y=count)) + geom_point() + geom_vline(xintercept="04_2023", color="red", linetype="dashed") + theme(axis.text.x=element_text(angle=90)) + 
  labs(title = "Rat Sightings Pre and Post Rat Czar",
       x = "Month-Year",
       y = "Number of Rat Sightings")
```

The red line indicates when the Rat Czar was appointed. We are interested in the slopes of the 6 months before and 6 months after:

```{r prereg, echo=FALSE, warning=FALSE}
# linear regression before and after
# 6 months before
czar_before = rat_czar |> filter(month_year %in% c("10_2022","11_2022", "12_2022", "01_2023", "02_2023", "03_2023", "04_2023")) |>
  mutate(months = case_when(month_year == "10_2022" ~ 1,
                           month_year =="11_2022" ~ 2,
                          month_year == "12_2022" ~ 3,
                         month_year ==  "01_2023" ~ 4,
                         month_year ==  "02_2023" ~ 5,
                          month_year == "03_2023" ~ 6,
                         month_year == "04_2023" ~ 7)) 
```

```{r before}
czar_before |> ggplot(aes(x=months, y=count)) + geom_point() + geom_line() + 
  labs(title = "Rat Sightings by Months (1=Oct 2022, 7=Apr 2023)",
       x = "Months",
       y = "Number of Rat Sightings")

czar_before |>
  lm(count~months, data=_) |> broom::tidy() |> knitr::kable()
```

The slope of rat sightings in the months pre-appointment is increasing, but not significant. What about the 6 months after?

```{r postreg, echo=FALSE, warning=FALSE}
czar_after = rat_czar |> filter(month_year %in% c("04_2023","05_2023","06_2023", "07_2023", "08_2023", "09_2023", "10_2023")) |>
  mutate(months = case_when(month_year == "04_2023" ~ 1,
                            month_year == "05_2023" ~ 2,
                           month_year =="06_2023" ~ 3,
                          month_year == "07_2023" ~ 4,
                         month_year ==  "08_2023" ~ 5,
                         month_year ==  "09_2023" ~ 6,
                          month_year == "10_2023" ~ 7)) 
```

```{r after}
czar_after |> ggplot(aes(x=months, y=count)) + geom_point() + geom_line() + 
  labs(title = "Rat Sightings by Months (1=Apr 2023, 7=Oct 2023)",
       x = "Months",
       y = "Number of Rat Sightings")

czar_after |>
  lm(count~months, data=_) |> broom::tidy() |> knitr::kable()
```

While we find that there are changes in the slope post-appointment, they are not statistically significant. As such, it seems like there are probably a number of other factors associated with rat sightings.

