---
title: "Data Visualization"
output: 
  html_document:
    toc: true
    toc_float: true
---

This page provides highlights of our analyses and includes what we believe to be our most important data plots. 

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = FALSE, message=FALSE, warning=FALSE)
library(tidyverse)
library(plotly)
library(lubridate)
library(dplyr)
library(choroplethr)
library(choroplethrMaps)
```

```{r rat sightings import}
sightings = read_csv('data/NYC_Rat_Sightings.csv') |>
  janitor::clean_names() |>
  separate(created_date, into=c("month","e", "day","f", "year", "g", "time"), sep=c(2,3,5,6,10,11)) |>
  select(-e,-f,-g) |>
  mutate(date = paste(year, month, day, sep=""),
         date = as.numeric(date)) |> 
  filter(date <= 20231031,
         date >= 20160101,
         !incident_zip <= 10000,
         !incident_zip > 11697,
         !borough %in% c("Unspecified", NA)) |>
  select(-agency, -agency_name, -complaint_type, -descriptor, -landmark, -facility_type, -park_facility_name, -vehicle_type, -taxi_company_borough, -taxi_pick_up_location, -bridge_highway_name, -road_ramp, -bridge_highway_segment, -bridge_highway_direction) |> select(unique_key, date, year, month, day, everything())
```

```{r weather import}
weather_df = rnoaa::meteo_pull_monitors(
    c("USW00094728"),
    var = c("PRCP", "TMIN", "TMAX"), 
    date_min = "2016-01-01",
    date_max = "2022-12-31") |>
  mutate(
    name = recode(id, USW00094728 = "CentralPark_NY"),
    tmin = tmin / 10,
    tmax = tmax / 10) |>
  select(name, id, everything())
```

```{r weather join}
sightings$date <- as.Date(as.character(sightings$date), format = "%Y%m%d")

rat_weather =
  right_join(sightings, weather_df, by="date")
```


## Overall Rat Sightings
First are two plots to show rat sightings over time and rat sightings by borough.

```{r timeseries}
rat_weather |>
  mutate(year = as.factor(year)) |>
  group_by(date) |>
  summarize(count = n(), year) |>  
  ggplot(aes(x = date, y = count, color=year)) +
  geom_point(alpha = 0.25) + 
  labs(title = "Time Series: Rat Sightings Over Time",
       x = "Date",
       y = "Number of Rat Sightings")
```

```{r borough}
# total count per year by borough - most in brooklyn and manhattan
sightings |> 
  group_by(year, borough) |> 
  summarize(count=n()) |> 
  plot_ly(x=~year, y=~count, color=~borough, type="scatter", mode="line") |> layout(title = "Rat Sightings by Borough", xaxis=list(title="Year"), yaxis=list(title="Number of Rat Sightings"))
```


## Rat Sightings & Weather
The following plot shows the trends of maximum temperature and number of rat sightings.

```{r temperature}
rat_weather |>
  group_by(date) |>
  summarize(count = n(), year, tmax) |>
  ggplot(aes(x = tmax, y = count, color = year)) + 
  geom_point(alpha = .3) +
  geom_smooth(se = FALSE, color="black") + 
  facet_grid(. ~ year) +
  labs(title = "Maximum Temperature vs. Rat Sightings by Year",
       x = "Maximum temperature (degrees Celsius)",
       y = "Number of Rat Sightings")
```

## Rat Sightings & Trash
```{r trash import and merge}
sightings = sightings |>
  select(unique_key, date, year, month, day, everything()) |> 
  mutate_all(tolower) |>
  mutate(across(where(is.character), trimws)) |> 
  group_by(year, month, borough) |> 
  summarize(ratcount = n()) 

waste <-
  read.csv("data/DSNY_Monthly_Tonnage_Data_20231202.csv") |> 
  janitor::clean_names() |> 
  separate(month, into = c("year", "month"), sep = " / ") |>
  group_by(year, month, borough) |>
  filter(year >= 2016 & year < 2024 ) |>
  summarize(
    total_refuse = sum(refusetonscollected, na.rm = TRUE),
    total_paper = sum(papertonscollected, na.rm = TRUE),
    total_mgp = sum(mgptonscollected, na.rm = TRUE))  |> 
  mutate_all(tolower) |> 
  mutate(across(where(is.character), trimws)) 

waste_2 = waste |> 
   pivot_longer(
    total_refuse:total_mgp, 
    names_to = "type",
    values_to = "tons") |> 
  mutate(type = substr(type, 7, 12),
         tons = as.numeric(tons)) |> 
  group_by(year, month, borough, type, tons)

merged = inner_join(waste_2, sightings, 
            by = c("year", "month", "borough")) |> 
  mutate(tons = as.numeric(tons)) |> 
  mutate(
    combined_ym = paste(year, month, sep = "-"),
    combined_ym = ym(combined_ym)) 

merged_tons = merged |> 
  group_by(year, month, borough, ratcount) |> 
  summarize(total_tons = (sum(tons)))  |> 
  mutate(
    combined_ym = paste(year, month, sep = "-"),
    combined_ym = ym(combined_ym) )
```

This plot shows the trash density by borough. 

```{r trash graph}
ggplot(merged_tons, aes(x=combined_ym, y = total_tons, color = borough, group = interaction(year, borough))) + 
    geom_line() + 
    labs(title = "Total Tons of Trash by Borough",
         x = "Date",
         y = "Total Tons",
         caption = "Data from DSNY Monthly Tonnage Data") 
```

## Rat Sightings & Restaurants

```{r inspections_df}
#Importing NYC Restaurant Inspections csv
inspections_df = read_csv("./data/NYC_Restaurant_Inspections.csv") |>
  janitor::clean_names() |>
  separate(inspection_date, c("Month", "Day", "Year"), sep = "/") |>
  select(-Day) |> #can remove other columns as needed
  filter(as.numeric(Year) > 2015)
```

```{r violation plot trend}
#Observations of violations
violations_plot_df <- inspections_df |>
  group_by(Year, boro, violation_code) |>
  mutate(code_obs = n())
```

This shows the top 20 restaurant cuisines with rat violations.

```{r top 20 cuisine 04k violations}
#Plotting 04K violations by cuisine type
violations_plot_df |>
  filter(violation_code == "04K") |>
  group_by(cuisine_description) |>
  mutate(cuisine_obs = n()) |>
  filter(cuisine_obs > 25) |>
  ggplot(aes(x = reorder(cuisine_description, cuisine_obs), 
             y = cuisine_obs)) +
  geom_point(aes(color = cuisine_description)) +
  theme(axis.text.x = element_text(angle = 40, hjust = 1)) +
  labs(title = "Top 20 Rat Violation Offenders by Cuisine Type 2016-2023") +
  theme(legend.position = 'none') +
  xlab("Cuisine Type") +
  ylab("Number of Rat Violations") +
  scale_y_continuous(breaks = c(50, 100, 150, 200, 250, 300))
```

## Rat Sightings & Socioeconomic Factors
```{r crowding data}
crowding18 =
  read_csv("data/crowding/2018_crowding.csv", col_names = TRUE) |> 
  janitor::clean_names() |> 
  select(geo_id, name, s2501_c01_006e, s2501_c01_007e, s2501_c01_008e, s2501_c01_001e) |> 
  rename(
    less_than_1_y18   = s2501_c01_006e,
    bet_1to1.5_y18    = s2501_c01_007e,
    more_than_1.5_y18 = s2501_c01_008e,
    total_homes_y18   = s2501_c01_001e,
  ) |> 
  drop_na() |> 
  filter(!row_number() %in% c(1)) |> 
  mutate_at(c("less_than_1_y18","bet_1to1.5_y18","more_than_1.5_y18", "total_homes_y18"), as.numeric) |> 
  mutate(
    less_prop_y18  = less_than_1_y18/total_homes_y18,
    bet_prop_y18  = bet_1to1.5_y18/total_homes_y18,
    more_prop_y18 = more_than_1.5_y18/total_homes_y18,
  )

crowding19 =
  read_csv("data/crowding/2019_crowding.csv", col_names = TRUE) |> 
  janitor::clean_names() |> 
  select(geo_id, s2501_c01_006e, s2501_c01_007e, s2501_c01_008e, s2501_c01_001e) |> 
  rename(
    less_than_1_y19   = s2501_c01_006e,
    bet_1to1.5_y19    = s2501_c01_007e,
    more_than_1.5_y19 = s2501_c01_008e,
    total_homes_y19   = s2501_c01_001e,
  ) |> 
  drop_na() |> 
  filter(!row_number() %in% c(1)) |> 
  mutate_at(c("less_than_1_y19","bet_1to1.5_y19","more_than_1.5_y19", "total_homes_y19"), as.numeric) |> 
  mutate(
    less_prop_y19  = less_than_1_y19/total_homes_y19,
    bet_prop_y19  = bet_1to1.5_y19/total_homes_y19,
    more_prop_y19 = more_than_1.5_y19/total_homes_y19,
  )

crowding20 =
  read_csv("data/crowding/2020_crowding.csv", col_names = TRUE) |> 
  janitor::clean_names() |> 
  select(geo_id, s2501_c01_006e, s2501_c01_007e, s2501_c01_008e, s2501_c01_001e) |> 
  rename(
    less_than_1_y20   = s2501_c01_006e,
    bet_1to1.5_y20    = s2501_c01_007e,
    more_than_1.5_y20 = s2501_c01_008e,
    total_homes_y20   = s2501_c01_001e,
  ) |> 
  drop_na() |> 
  filter(!row_number() %in% c(1)) |> 
  mutate_at(c("less_than_1_y20","bet_1to1.5_y20","more_than_1.5_y20", "total_homes_y20"), as.numeric) |> 
  mutate(
    less_prop_y20  = less_than_1_y20/total_homes_y20,
    bet_prop_y20  = bet_1to1.5_y20/total_homes_y20,
    more_prop_y20 = more_than_1.5_y20/total_homes_y20,
  )
  
crowding21 =
  read_csv("data/crowding/2021_crowding.csv", col_names = TRUE) |> 
  janitor::clean_names() |> 
  select(geo_id, s2501_c01_006e, s2501_c01_007e, s2501_c01_008e, s2501_c01_001e) |> 
  rename(
    less_than_1_y21   = s2501_c01_006e,
    bet_1to1.5_y21    = s2501_c01_007e,
    more_than_1.5_y21 = s2501_c01_008e,
    total_homes_y21   = s2501_c01_001e,
  ) |> 
  drop_na() |> 
  filter(!row_number() %in% c(1)) |> 
  mutate_at(c("less_than_1_y21","bet_1to1.5_y21","more_than_1.5_y21", "total_homes_y21"), as.numeric) |> 
  mutate(
    less_prop_y21  = less_than_1_y21/total_homes_y21,
    bet_prop_y21  = bet_1to1.5_y21/total_homes_y21,
    more_prop_y21 = more_than_1.5_y21/total_homes_y21,
  )

crowding_10s = 
  full_join(crowding18, crowding19, by = "geo_id")

crowding_20s = 
  full_join(crowding20, crowding21, by = "geo_id")

crowding_all =
  full_join(crowding_10s, crowding_20s, by = "geo_id") |> 
  drop_na() |> 
  mutate(
    id = str_sub(geo_id, 10),
    county = str_extract(name, "(?<=,)[^,]+(?=,)")
  ) |> 
  filter(county %in% c(" Kings County"," Bronx County"," Queens County"," Richmond County",
                       " New York County"))
```


Finally, we compared crowding and rat sightings and provided a chloropeth map of the Bronx here.
```{r ses sightings}
sightings = read_csv('data/NYC_Rat_Sightings.csv') |> 
  janitor::clean_names() |> 
  separate(created_date, into=c("month","e","day","f","year","g","time"), sep=c(2,3,5,6,10,11)) |> 
  select(-e,-f,-g) |> 
  mutate(date = paste(year, month, day, sep=""),
         date = as.numeric(date)) |>  
  filter(date <= 20231031, date >= 20160101, !incident_zip <= 10000, !incident_zip >11697, !borough %in%
           c("Unspecified", NA)) |> 
  select(-agency, -agency_name, -complaint_type, -descriptor, -landmark, -facility_type, -park_facility_name,
         -vehicle_type, -taxi_company_borough, -taxi_pick_up_location, -bridge_highway_name, -road_ramp,
         -bridge_highway_segment, -bridge_highway_direction) |> 
  select(unique_key, date, year, month, day, everything()) 

crowding_Bronx = crowding_all |> 
  pivot_longer(less_than_1_y18:more_prop_y21,
               names_to = "occupancy",
               values_to = "count") |> 
  mutate(year     = str_sub(occupancy, -3),
         category = str_sub(occupancy, end=-5)) |> 
  mutate(
    year = case_match(
      year,
      "y18" ~ 2018,
      "y19" ~ 2019,
      "y20" ~ 2020,
      "y21" ~ 2021
    )
  ) |> 
  filter(category == "more_prop",
         year == "2021",
         county == " Bronx County") |> 
  rename(region = "id",
         value  = "count") |> 
  select(region, value, county) |> 
  mutate_at(c("region"), as.numeric)

```

```{r bronx sightings}
sightings |>
  filter(year == "2021",
         city == "BRONX") |>
  plot_ly(x=~longitude, 
          y=~latitude, 
          type="scatter", 
          mode="markers", 
          marker=list(color="darkblue",size=2,opacity=0.5)) |> 
  layout(title = "Rat Sightings in the Bronx", xaxis=list(title="Longitude"), yaxis=list(title="Latitude")) 
```

```{r bronx chloro, results='hide'}

tract_choropleth(crowding_Bronx,"new york", county_zoom = 36005,
                                         title = "2021 Proportion of Households with 1.51 or More Occupants/Room in 
                                         Bronx by Census Tract",
                                         legend = "Proportion")
```
